Constraining accessibility and the family of spatial interaction models" #"A family of accessibility measures (bringing balance back?)

### 2025-02-04 paper/manuscript.qmd
Finished good-enough draft up until (and include) sub-section "##Re-interpreting SIM constraints for accessibility"

---
title: "Accessibility, constraints and the family of spatial interaction models" #"A family of accessibility measures (bringing balance back?)"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G. Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959]. In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction" [p.73], as "a generalization of the population-over-distance relationship or 'population potential' concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73]. As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads, and likely preoccupied with applied research into the power of transportation to shape land uses. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind. @hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns . In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
$$
S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
$${#eq-accessibility}

@@

## Gravity-based accessibility's cousin: Wilson's spatial interaction model
<!--SUMMARY: State how accessibility has the same Newtonian roots as Wilsons' family of spatial interaction models. Introduce SIM's formulation. Emphasis the interpretation that comes from $k$. Emphasis how while SIM has Newtonian roots, its proportionality constant(s) are actually balancing factors derived via entropy-maximisation, obtaining the interaction variable as a statistical average. -->
Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used today in transportation demand modelling [@ortuzar_2011_modelling;@wilson1971]. However, in many ways these spatial interaction models (AKA gravity models) continued to evolve independently from accessibility (@eq-accessibility-general) as will be discussed next section. Our paper aims to demonstrate that Hansen-type accessibility can be interpreted as a SIM. This connection can enrich the accessibility literature by linking it to the extensive practical application of SIMs. This paper's focus is on how SIMs utilize "proportionality constants" and constraints to interpret results, and how Hansen-type accessibility can transit to _constrained_ accessibility. 
To begin, the SIM as defined in Wilson [-@wilson1971], takes the following form: 
$$
T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
$${#eq-phys-gravity-model}

@@

The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!--NOTE: in Wilson's entropy-max derivation, this is actually eq-constraint0_5-gravitymodel:-->
$$
\sum_i\sum_j T_{ij} = T
$${#eq-constraint0-gravitymodel}

@@

Using these constraints, the SIM @eq-phys-gravity-model is distinguished into four general cases. These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling]:

- The unconstrained case: neither @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
- The total constrained case: only @eq-constraint0-gravitymodel hold <!--NOTE: not sure if this is consistent with entropy-maximisation?-->
- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->

In each of these four cases, the single proportionality constant $k$ is replaced by balancing factors $A_i$, $B_j$, and $K$ as follows to ensure the case-specific constraints are satisfied:
$$
A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}
$${#eq-bf-A-gravitymodel}

@@

For the doubly-constrained case, $A_i$ and $B_j$ must be solved together along with $K$. For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel. For both singly-constrained cases, $K$ also holds. For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint. And for the unconstrained case, all other cases (the total constrained, singly-constrained and doubly-constrained cases) are simply constrained variations.

## Background on accessibilities' distinct development away from spatial interaction models
<!--COME-BACK HERE-->
<!--SUMMARY: Accessibility developed separate from SIM.

@@

The institutionalization of research on spatial interaction modelling was fueled by advancements in computer technologies and the substantive and unprecedented Federal-Aid Highway Act of 1956, which authorized $24.8 billion in spending at a 90% federal share from between 1957 to 1969 to begin and complete the US Interstate Highway System [@weinerUrbanTransportationPlanning2016]. This ambitious project went on to shape transportation planning practice globally []. Within the same time period, accessibility measures were not applied practically at such a large scale. Instead, accessibility measures were taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners; remaining more in theory than in practice.  However, due to growing critiques about traditional transportation modelling practice focus on  _mobility_, calls for a shift to accessibility are intensifying [@handy2020] and accessibility planning is beginning to be present in practice [] <!-- citation on where they are applied in practice? US guidelines. Netherlands. -->

Despite the theoretical and observed reality that both mobility and accessibility influence an individual’s capacity to travel [@morrisAccessibilityIndicatorsTransport1979], SIM literature that focuses on Hansen's accessibility is relatively scarce in the literature. To capture this sentiment in the literature, @fig-Fig1 demonstrates the bibliometric coupling of papers that cites @hansen1959, @wilson1971, and both @hansen1959 and @wilson1971. @fig-Fig1 is based on a SCOPUS search of papers that cite @hansen1959's DOI (2,134 documents) and another SCOPUS search of papers that cite @wilson1971's DOI (368 documents) conducted on November 9, 2024 and a bibliometric coupling matrix created using the {bibliometrix} R package [@bibliometrix2017].

@@

Many of the papers that cite both @hansen1959 and @wilson1971 (77 journal articles) are transportation modelling papers working on integrating accessibility within the SIM or interpreting HAM within the SIM. For instance, @battyMethodResiduesUrban1976 states that the relationship between the SIM and HAM may be worth exploring. There is also a consistent acknowledgement in the literature that HAM and SIM are related. Specifically, that HAM are within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985]. Recognizing the balancing factors as HAM is a "common sense" approach at integrating accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979].

There are other approaches to consider accessibility with SIM beyond HAM which can be founded on micro-economic consumer behaviour [@morrisAccessibilityIndicatorsTransport1979]. For instance, others explore entropy maximizing solutions (Wilson's framework and hence the balancing factors defined as Hansen's accessibility) when considering demand behavior described by random utility choice models [@leonardiRandomUtilityDemand1984].

Some work on location-allocation models also used both Hansen and Wilson. For instance, @leonardiOptimumFacilityLocation1978 and @beaumontLocationallocationProblemsPlane1981 define a location-allocation problem using the singly-constrained SIM, and then defines one using Hansen-type accessibility as the objective function subject.

Then, @fotheringhamSpatialCompetitionAgglomeration1985 demonstrated that destination's accessibility should be explicitly specified within Wilson's framework, citing Hansen measure as convenient formulation. Without the consideration for accessibility explicitly, unexplained spatial pattern remains. This later came to be the 'competition destination' model, and is one of a few techniques to address spatial variation in SIMs. [@oshanSpatialStructureDebate2021]. Implicitly, we gather that at this point: it was understood balancing factors in the SIM could not be understood as capturing accessibility. We also gather that in this moment of growing complexity of spatial interaction modelling, accessibility branched off and sprouted in the medium of budding GIS technology and the emergence of space-time dimensions [@millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting]. With the possibility for showcasing complex GIS methodologies for varied case studies, accessibility became mature enough to branch off on its own.

In papers after @fotheringhamSpatialCompetitionAgglomeration1985, Hansen is typically cited as providing a definition of accessibility but how Wilson is cited changes through the years. Wilson is attributed for using different travel cost functions in depending on the context [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].

Othertimes, Wilson and Hansen measures were progressively incorrectly conflated throughout the years [ @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring] - indeed they both measure spatial interaction so are somewhat related, but they are not the same. They are also co-cited to refer to the concept of 'gravity' in models [@liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]

And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. This interpretation is fraut, upon considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration, which many of these works do not contend with.

THE SPATIAL STRUCTURE DEBATE: @oshanSpatialStructureDebate2021

Some papers also use SIMs to estimate trip or some other interaction flows alongside accessibility indicators [@clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].

Hansen is a relatively straightforward simple application, while Wilon's contribution is methodological and dense - must be carefully read. Many takeaways can be drawn, so with accessibilities changing applications throughout the years -- different things about Wilon's contribution becomes relevant. The most contemprary connection between the two is in the relam of competitive accessibility -- in the interpretation of Wilson's balancing factors as a way to consider competition. We fit this contribution in this literature. Competition fits on a spectrum - by introducing balancing factors trips are _constrained_ to equal regional or zonal totals. This same application can be applied to accessibility.

Some 
Additionally, trending research topics of the time such as space-time and more  []

previously mentioned in accessibility development: the breadth of opportunity $O_j$ that can be studied [] as well as the accuracy of parameters that capture distance-decay $f(c_{ij})$ []. Outside of these factors, accessibility literature that delves into competition, does not cite Wilson, with some exception... Soukhov for the balancing factor.

Overall, work that cities Wilson is closer related to transportation demand modelling than accessibility. However there are a few works of the 77 documents that are in the realm of accessibility. ...

However there are a few papers that are accessibility focused that focus on

<!-- NOTE: you can validate SIM e.g., how well does the model predict a known Tij matrix? but what is a known Vij matrix, potential accessible opportunities at i for j? Harder to say. -->

There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models.

## Defining accessibility constraints (a potential spatial interaction model?)

@@

Example of solving the population-constrained $V_i''^{(1)}$:
<!--COMEBACKHERE-->

@@

>>>>

---
title: "Constraining accessibility and the family of spatial interaction models" #"A family of accessibility measures (bringing balance back?)"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G. Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959]. In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction... a generalization of the population-over-distance relationship or _population potential_ concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73]. As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads, and likely preoccupied with applied research into the power of transportation to shape land uses. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind. @hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns . In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
$$
S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
$${#eq-accessibility}

@@

## Gravity-based accessibility's cousin: Wilson's spatial interaction model
<!--SUMMARY: Accessibility has the same Newtonian roots as Wilsons' family of spatial interaction models. Introduce SIM's formulation. Emphasize the interpretation that comes from $k$. While SIM has Newtonian roots, its proportionality constant(s) are actually balancing factors derived via entropy-maximisation, obtaining the interaction variable as a statistical average. -->
Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used today in transportation demand modelling [@ortuzar_2011_modelling;@wilson1971]. However, in many ways these spatial interaction models, commonly refered to as gravity models, continued to evolve independently from accessibility (@eq-accessibility-general). Our paper aims to demonstrate that Hansen-type accessibility can be interpreted as a SIM. This connection can enrich the accessibility literature by linking it to the extensive practical application of SIMs. This paper's focus is on how SIMs utilize "proportionality constants" and constraints to interpret results, and how Hansen-type accessibility can transit to _constrained_ accessibility.

To begin, the SIM as defined in Wilson [-@wilson1971], takes the following form: 
$$
T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
$${#eq-phys-gravity-model}

@@

The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!--NOTE: in Wilson's entropy-max derivation, this is actually eq-constraint0_5-gravitymodel a TOTAL TRAVEL COST constraint:-->
$$
\sum_i\sum_j T_{ij} = T
$${#eq-constraint0-gravitymodel}

@@

Using these constraints, the SIM @eq-phys-gravity-model is distinguished into three general cases. These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling]:

- The unconstrained case: neither @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold, only @eq-constraint0-gravitymodel - the total number of trips, is used;
- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->

In each of these cases, the single proportionality constant $k$ is replaced by balancing factors $A_i$, $B_j$, and $K$ as follows to ensure the case-specific constraints are satisfied:
$$
A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}
$${#eq-bf-A-gravitymodel}

@@

For the doubly-constrained case, $A_i$ and $B_j$ must be solved together along with $K$. For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel. For both singly-constrained cases, $K$ also holds. For the unconstrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.

These constraints (@eq-bf-A-gravitymodel, @eq-bf-A-gravitymodel, and @eq-bf-K-gravitymodel) and the SIM @eq-phys-gravity-model can be expressed not just for the variable of _interaction_ (e.g., the number of trips) but also as a variable of _potential interaction_. Before detailing SIM's formulation for accessibility, however, background on the development of accessibility and SIM independent from eachother is explained.

## Background on accessibilities' distinct development from spatial interaction models
<!--
NOTE: This section is a tough one... please suggest ideas...

SUMMARY: Accessibility developed separate from SIM.

@@

Current transportation planning practice was arguably born in the 1950s; fueled by advancements in computer technologies and catalyzed by the unprecedented Federal-Aid Highway Act of 1956 which authorized the creation of the US Interstate Highway System, costing over a hundred billion in public spending in the following 35 years [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007]. To plan this ambitious project, transportation planning practice adopted SIM and extended it into four-step transportation demand modelling; where tangible outputs of transportation systems such as _mobility_ (e.g., trips) were calibrated, modeled and applied to "predict and provide" transportation supply and demand [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016]. However, within the same time period, accessibility remained largely within the academic realm. For instance, accessibility measures were used to study a variety of transportation-related dimensions such as residential (or firm) location and city growth [e.g., @kain1962multiple]; where the importance of accessibility versus non-accessibility characteristics (quality of destination, composition of household or inherent individual characteristics) were debated [e.g., @redding1970quality; @chapin1968activity] and experimentation with different cases and travel behaviours emerged showing this nuance [e.g., <!--studies that showcase nuanced preferences in accessibility based on household characteristics or destination type etc.-->]. We perceive these debates as a significant point that defined accessibility's outgrowth from the mobility-oriented practical implementation of SIM in transportation planning. Accessibility's flexible definition, various mathematical formulations and increasing complexity in its application and interpretation made it a concept more difficult to institutionalize in the moments were planning "embraced" automobility-oriented regional development.

However, in recent years, this embrace has been weakened by the mounting evidence of the burdens that automobility is placing on the planet and communities. A renewed wave of criticism levelled at traditional transportation modelling practice's focus on  _mobility_ instead of _accessibility_ have been growing [@handy2020]. Accessibility, captures the concept of _potential_ interaction, more aligned with the true purpose of transportation systems, and can be used to plan for equitable and sustainable transportation systems in ways using _mobility_ cannot. While SIM and mobility-based planning has been institutionalized in transportation practice, accessibility has yet to do so [@papaAccessibilityInstrumentsPlanning2015; @boisjolyInsiderPlannersPerspective2017; @silvaProximitycentredAccessibilityConceptual2023].

To capture this sentiment in the literature, @fig-Fig1 demonstrates the bibliometric coupling of papers that cites @hansen1959 (one of accessibility's "seminal" works), @wilson1971 (one of SIM's "seminal" works), and both @hansen1959 and @wilson1971. @fig-Fig1 is based on a SCOPUS search of papers that cite @hansen1959's DOI (2,134 documents) and another SCOPUS search of papers that cite @wilson1971's DOI (368 documents) conducted on November 9, 2024 and a bibliometric coupling matrix created using the {bibliometrix} R package [@bibliometrix2017].

@@

In summary, papers that cite only @hansen1959 do not focus on SIM and instead focus on _potential_ for interaction while papers that cite only @wilson1971 deal with SIM in the realm of _interaction_. Those that cite both @hansen1959 and @wilson1971 (77 journal articles) are of interest in this paper, as they often focus on accessibility's role within the SIM formulation (@eq-phys-gravity-model) but are often still seem to be more SIM than HAM, hence are overviewed as follows.

Often, these SIM/HAM papers seek interpretation of the HAM within the SIM. Specifically, many early works in the 70s and 80s have indicated that HAM is represented within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985]. Recognizing the balancing factors as HAM is a "common sense" approach at including accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979], though its relationship may be worth further exploring [@battyMethodResiduesUrban1976].

However, some authors that did explore this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how SIM may insufficiently explain spatial patterns and how explicitly defining destination's accessibility, citing HAM as convenvient formulation, within the SIM framework may remedy the issue. This re-formulation of the SIM later came to be defined as _competition destination_ model, and is now one of a few techniques to address spatial variation in SIMs [@oshanSpatialStructureDebate2021]. Other early SIM works incorporating accessibility also went implicitly beyond this understanding that accessibility is neatly considered within the SIM framework. Some works went beyond HAM and considered other accessibility expressions of individual choice. For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984]. Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly-constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].

Interestingly from this work in the 70s and 80s, it can be understood that while the balancing factors within the SIM could be read as HAM, they should not be understood as capturing accessibility <!--eek? I don't think I can say this-->. We also gather that in this moment of growing complexity of SIM (not) capturing accessibility, literature exploring accessibility sprouted in new mediums of budding GIS technologies and the conceptual emergence of space-time dimensions in the 90s onward [e.g., @millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting]. With the possibility for showcasing complex GIS methodologies for varied case studies, this accessibility-based literature became mature enough to branch off on its own without meting in the affairs of SIM.

In papers published within and after the 1990s, @hansen1959 is consistently cited as providing a definition of accessibility or in using the HAM, however the attribution of @wilson1971 is various. @wilson1971 is cited alongside accessibility-based literature in three important ways; two of which incorrectly conflate and interpret the SIM and HAM.

Firstly, many of these works attribute @wilson1971 for using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].

Secondly, in the 2000s and beyond, the concept of spatial interaction and accessibility are conflated throughout the years [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]; indeed, accessibility is an expression of _potential_ spatial interaction, and should not be interpreted the same way as the SIM is commonly used. Surprisingly, @hansen1959 and @wilson1971 are further conflated by being co-cited as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].

And thirdly, accessibility-based literature cite @wilson1971's balancing factors, and re-iterate how they can be interpreted as the inverse of HAM and hence can capture competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. However, we argue that this interpretation is fraut, considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration that SIM does not completely incorporate accessibility. However some work does consider competition more scrupulously; these works use SIMs to estimate trip (or some other interaction flows) alongside accessibility indicators [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating]. Only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.

Indeed, @hansen1959 offers a relatively straightforward and practical application of accessibility, while @wilson1971's contribution is methodological and dense. Accessibility-based literature that has flourished since the 1990s, sprouting from the rise of GIS technologies, the plethora of _bigger_ data and different qualities of opportunities and populations to examine, it has developed in a different direction. However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches. A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interation_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors. How these balancing factors are applied in the context of _interaction_ like trips  can be applied to accessibility outputs like the "number of opportunities potentially interacted with".

## Re-interpreting SIM constraints for accessibility

@@

Example of solving the population-constrained $V_i''^{(1)}$:
<!-- COMEBACKHERE -->

@@

<!-- Despite the theoretical and observed findings that _both_ mobility and accessibility influence an individual’s capacity to travel [@morrisAccessibilityIndicatorsTransport1979], SIM literature that focuses on Hansen's accessibility is relatively scarce in the literature.  -->
